---
format:
  html:
    code-copy: true
    code-fold: false
    code-tools: false
execute:
  eval: false
  echo: fenced
---

# Figures, tables, and cross-references

To maintain professional and navigable documents, all key results must be appropriately labeled and referenced within the text.

## Figures

For a chunk to be treated as a cross-referenceable figure, its label **must** start with the fig- prefix. You must also provide a fig-cap.

```{r}
#| label: fig-map
#| fig-cap: "Spatial distribution of marine protected areas."
#| eval: true
mtcars |>
  dplyr::mutate(cyl = factor(cyl)) |>
  ggplot2::ggplot(ggplot2::aes(x = wt, y = mpg, color = cyl)) +
  ggplot2::geom_point(size = 2.5, alpha = 0.8) +
  ggplot2::labs(
    title = "Fuel economy decreases as car weight increases",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon",
    color = "Cylinders"
  )
```

## Tables

Table labels **must** start with the tbl- prefix. At emLab, we primarily use three packages for table generation, depending on the complexity required:

### knitr::kable

Best for simple, clean tables with minimal formatting needs. It is lightweight and works perfectly with Quarto’s internal styling.

### kableExtra

Use this when you need complex formatting, such as grouping rows, adding footnotes, or specific cell styling for PDF/HTML.

### gt (Grammar of Tables)

The preferred modern choice for highly customized HTML tables. It provides a "ggplot-like" syntax for building tables piece-by-piece.

## Exporting tables to LaTeX (.tex)

If you are collaborating on a paper in Overleaf or a standalone LaTeX document, you should export your tables to .tex files.

**Using kableExtra:**

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics of model performance."
knitr::kable(summary_df, format = "latex", booktabs = TRUE) |>
  save_kable("tables/summary_table.tex")
```

**Using gt:**

```{r}
gt_table |>
  gtsave("tables/summary_table.tex")
```

## Referencing in text

Reference these elements in your narrative using the @ symbol followed by the label.

- "As shown in the figure labeled `fig-map` (@fig-map), the distribution is skewed..."
- "The results in the table labeled `tbl-summary` (@tbl-summary) indicate a significant trend..."
