---
format:
  html:
    code-copy: true
    code-fold: false
    code-tools: false
execute:
  eval: false
  echo: fenced
---

# Integrating with a targets pipeline

For projects using the `targets` package, rendering the Quarto document should be the final step in the pipeline. We recommend using `tarchetypes::tar_quarto()` for this rather than wrapping a render call in a plain `targets::tar_target()`.

## Why `tarchetypes::tar_quarto()` instead of `targets::tar_target()`?

You could imagine that one approach to render a quarto document in a targets pipeline would be to define a target like this:

```{r}
# Not recommended
targets::tar_target(report, quarto::quarto_render("analysis/final_report.qmd"))
```

However, the problem is that `targets` has no visibility into what the `.qmd` file actually depends on. It treats the render as a black box, so it cannot detect when the document's upstream targets have changed and the report needs to be re-rendered. You end up either re-rendering on every `targets::tar_make()` call or, worse, serving a stale report without realizing it.

`tarchetypes::tar_quarto()` solves this by analyzing your `.qmd` file for `targets::tar_read()` and `targets::tar_load()` calls and automatically registering those targets as dependencies. This means `targets` actually re-render the report when something the quarto document actually reads from the pipeline has changed — no manual dependency tracking required.

```{r}
# Recommended
tarchetypes::tar_quarto(
  name = quarto_notebook,
  path = "qmd/quarto_notebook.qmd",
  quiet = FALSE
)
```

Inside your `.qmd` file, use `targets::tar_read(target_name)` or `targets::tar_load(target_name)`to pull results from the pipeline. These calls are what `tar_quarto()` scans to build the dependency graph.

If you have a strong reason to use `targets::tar_target()` directly (e.g., a non-standard render workflow), it is still possible, but you will need to manually declare dependencies using inside the target expression to ensure correct invalidation behavior. We strongly recommend simply using `tarchetypes::tar_quarto` instead.
