# Code

All data processing and analysis should be performed with code (i.e., avoid spreadsheets), and all code should be packaged in scripts that are version controlled and follow a style guide. Using code and scripts allows for better organization, documentation, and reproducibility of analysis workflows. **All emLab code should be stored in the emLab GitHub account.**

**Recommended readings:**

* Bryan, Jennifer. 2018. "Excuse Me, Do You Have a Moment to Talk About Version Control?" *The American Statistician* 72 (1): 20–27. <https://doi.org/10.1080/00031305.2017.1399928>.

* Stodden, Victoria, and Sheila Miguez. 2014. "Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research." *Journal of Open Research Software* 2 (1): e21. <https://doi.org/10.5334/jors.ay>.

* Wilson, Greg, D. A. Aruliah, C. Titus Brown, Neil P. Chue Hong, Matt Davis, Richard T. Guy, Steven H. D. Haddock, et al. 2014. "Best Practices for Scientific Computing." *PLOS Biology* 12 (1): e1001745. <https://doi.org/10.1371/journal.pbio.1001745>.

* Wilson, Greg, Jennifer Bryan, Karen Cranston, Justin Kitzes, Lex Nederbragt, and Tracy K. Teal. 2017. "Good Enough Practices in Scientific Computing." *PLOS Computational Biology* 13 (6): e1005510. <https://doi.org/10.1371/journal.pcbi.1005510>.


## Scripts and Version Control {#scripts-and-version-control}

Code should be crafted according to the following guidelines:

* Use scripts
* Document scripts, but not too much
* Organize scripts consistently (see format below)
* Use Git to version control scripts
* Make atomic Git commits (see description below)

Script files should be documented and organized in such a way to enhance readability and comprehension. For example, use a standardized header for general documentation and sections to make it easier to understand and find specific code of interest. Code should also be self-documenting as much as possible. Additionally, use relative filepaths for importing and exporting objects. Scripts should also be modular by focusing on one general task. For example, use one script for cleaning data, another script for visualizing data, etc. A makefile can then be used to document the analysis workflow. There is an art to this organization, so just keep in mind the general principle of making code easy to understand, for your future self and for others.

Here is an example template for R scripts:

```
# =============================================================================
# Name:           script.R
# Description:    Visualizes data
#
# Inputs:         data.csv
# Outputs:        graph.png
#
# Notes:          - Use a diverging color palette for best results
#                 - Output format can be changed as needed
# =============================================================================


# Set up environment ----------------------------------------------------------

library(tidyverse)

# Set path for Google Drive filestream based on OS type
# Note - this will work for Windows and Mac machines.
# If you use Linux, you will need to set your own path to where Google Drive filestream lives.

team_path <- ifelse(Sys.info()["sysname"]=="Windows","G:/","/Volumes/GoogleDrive/")

# Next, set the path for data directory based on whether project is current or archived.
# Note that if you use a different Shared Drive file structure than the one recommended in the "File Structure" section, you will need to manually define your data path.
# You should always double-check the automatically generated paths in order to ensure they point to the correct directory.

# First, set the name of your project

project_name <- "my-project"

# This will automatically determine if the project exists in the "current-projects" or "archived-projects" Shared Drive folder, and set the appropriate data path accordingly.
data_path <- ifelse(dir.exists(paste0(team_path,"Shared drives/emlab/projects/current-projects/",project_name)),
                       paste0(team_path,"Shared drives/emlab/projects/current-projects/",project_name,"/data/"),
                       (paste0(team_path,"Shared drives/emlab/projects/archived-projects/",project_name,"/data/")))


# Import data -----------------------------------------------------------------

# Load data from Shared Drive using appropriate data path
my_raw_data <- read_csv(paste0(data_path,"raw/my_raw_data.csv"))

# Process data ----------------------------------------------------------------


# Analyze data ----------------------------------------------------------------


# Visualize results -----------------------------------------------------------


# Save results ----------------------------------------------------------------


```

[Git](https://git-scm.com/) tracks changes in code line-by-line with the use of commits. Commits should be atomic by documenting single, specific changes in code as opposed to multiple, unrelated changes. Atomic commits can be small or large depending on the change being made, and they enable easier code review and reversion. Git commit messages should be informative and follow a certain style, such as the guide found [here](https://chris.beams.io/posts/git-commit/). There is also an art to the version control process, so just keep in mind the general principle of making atomic commits.

More advanced workflows for using Git and GitHub, such as using pull requests or branches, will vary from project to project. It is important that the members of each project agree to and follow a specific workflow to ensure that collaboration is effective and efficient.

## Code Styling

### The Philosophy of Style

At emLab, we view code not just as a set of instructions for a computer, but as a medium of communication between researchers. In a collaborative environment where projects span years and multiple contributors, **style is a form of documentation.**  
The primary goals of our code style guide are:

1. **Reduced Cognitive Load:** When every script follows the same conventions, your brain doesn't have to "re-parse" the structure of a file every time you switch projects.  
2. **Meaningful Diffs:** Standardized formatting ensures that GitHub "diffs" reflect actual logic changes rather than trivial changes in whitespace or indentation.  
3. **Error Prevention:** Many style rules (like those enforced by linters) catch common bugs—such as unused variables or misplaced parentheses—before the code is even run.  
4. **Inclusivity:** Clean, well-spaced code is more accessible to junior researchers and external collaborators.

### Style Guides by Language

We maintain specific style preferences for each language used at the lab. While these guides provide the rules, the [Code Formatters](#bookmark=id.rq10khlw8of2) section below explains how to automate them.

#### R and the Tidyverse

Our R code follows the [Tidyverse Style Guide](https://style.tidyverse.org/). While you should read the full guide, here are the non-negotiables for emLab projects:

* **Object names:** Use snake\_case for variables and functions (e.g., `sea\_surface\_temp`, not `SeaSurfaceTemp` or `sst`).  
* **Spacing:** Always put a space after a comma, and around most binary operators (`==`, `\+`, `\-`, `\<-`, etc.).  
* **Pipes:** Use the native pipe `|\>` (R 4.1+) or the magrittr pipe `%\>%`. Always put a space before the pipe and a new line after it.  
* **Indentation:** Use 2 spaces for indentation. Never use tabs.  
* **Long lines:** Limit code to 80 characters per line. This ensures scripts are readable on smaller laptop screens and in side-by-side GitHub views.

#### Other Languages

For other languages, we defer to established industry standards:

* **Python:** We follow [PEP 8](https://peps.python.org/pep-0008/), which is the de facto standard for Python code.  
* **Stata:** We follow the principles outlined in the [Stata Guide](https://www.stata.com/manuals/g.pdf) and common research best practices.  
* **General (C++, Java, etc.):** For languages not explicitly covered, we follow the [Google Style Guides](https://google.github.io/styleguide/).  
* **SQL:** Use uppercase for keywords (e.g., SELECT, FROM, WHERE) and snake\_case for table and column names.

### Code Formatters

A **code formatter** is a tool that automatically adjusts your code's layout—spacing, indentation, line breaks, and alignment—to match a specific style guide. Instead of manually fixing every comma or bracket, the formatter does it for you instantly.  
At emLab, we use a new generation of "Rust-based" tools. They are significantly faster than older tools and can be configured to run automatically every time you save a file.  
**Note:** To maintain consistency across the lab, the use of these formatters is **required** for all new code development.

#### R Formatting: Air (Required for R users)

**Air** is the required high-performance R formatter for emLab projects. It is developed by the Posit team and ensures strict adherence to Tidyverse standards.

##### Install the Extension

* In Positron or VS Code, open the Extensions Marketplace (`Cmd+Shift+X`).  
* Search for and install **"Air"** (published by Posit).

##### Enable "Format on Save"

To make formatting automatic, open your settings.json (`Cmd+Shift+P` \-\> search for "Open User Settings (JSON)") and paste the following:  

```r
{  
  "\[r\]": {  
    "editor.defaultFormatter": "posit.air",  
    "editor.formatOnSave": true  
  }  
}
```

#### Python Formatting: Ruff (Required for Python users)

**Ruff** is the required formatter and linter for all Python development at emLab. It is extremely fast and replaces several older tools with a single extension.

##### Install the Extension

* In Positron or VS Code, open the Extensions Marketplace (`Cmd+Shift+X`).  
* Search for and install **"Ruff"** (published by Astral Software).

##### Enable "Format on Save"

Open your settings.json and paste the following:  

```r
{  
  "\[python\]": {  
    "editor.defaultFormatter": "charliermarsh.ruff",  
    "editor.formatOnSave": true,  
    "editor.codeActionsOnSave": {  
      "source.fixAll.ruff": "explicit",  
      "source.organizeImports.ruff": "explicit"  
    }  
  }  
}
```

### RStudio Code Formatting

If you are using RStudio (where Air is not yet natively supported), you are still required to maintain styling using the `styler` package:

1. Install the package: install.packages("styler").  
2. Use the **Addins** menu in the top toolbar to select "Style active file" before every commit.  
3. Navigate to **Tools \-\> Global Options \-\> Code \-\> Saving** and check **"Ensure that source files end with a newline"**.

## Reproducibility

Prioritizing reproducibility when writing code code not only fosters collaboration with others during a project, but also makes it easier for users in the future (including yourself!) to make changes and rerun analyses as new data become available. Some useful tools and practices include:

- **Commenting code:** Adding brief but detailed comments to your scripts that document what your code does and how it works will help others understand and use your scripts. At the top of your scripts, describe the purpose of the code as well as its necessary inputs, required packages, and outputs.
- **File paths:** Avoid writing file paths that only work on your computer. Where possible, use relative file paths instead of absolute files paths so your code can be run by different users or operating systems. For R users, using [R Projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) and/or the [here](https://github.com/jennybc/here_here) package are great ways to help implement this practice.
- **Functions:** If you find yourself copying and pasting similar blocks of code over and over to repeat tasks, turn it into a function!
- **R Markdown:** R users can take advantage of [R  Markdown](https://rmarkdown.rstudio.com/lesson-1.html) for a coding format that seamlessly integrates sections of text alongside code chunks. R Markdown also enables you to transform your code and text into report-friendly formats such as PDFs or Word documents.
- **Git and GitHub:** [As described above,](#scripts-and-version-control) Git tracks changes to files line-by-line in commits that are attributable to each team member working on the project. GitHub then compiles the history of any Git-tracked file online, and synchronizes the work of all collaborators in a central repository. Using Git and GitHub allow multiple people to code collaboratively, examine changes as they occur, and restore prior versions of files if necessary.

We also recommend managing package dependencies and creating a reproducible coding pipeline. These two aspects are discussed in more detail below.

## Code Documentation

### Philosophy: Documentation as a Research Asset
At emLab, we view code documentation not as an afterthought, but as a core component of our research's reproducibility and longevity. High-quality documentation ensures that our "future selves" and external collaborators can understand the *intent* behind the analysis, not just the mechanics of the code.

*Core Principles:*

* **Code is for humans:** Write code that is easy for a human to read, using documentation to bridge the gap between "what" the code does and "why" it does it.
* **Self-documenting code first:** Strive for expressive variable and function names (e.g., `calculate_fishing_mortality()` rather than `f_mort_calc()`). Documentation should supplement, not fix, obscure code.
* **Documentation is near the code:** Keep documentation in the source files (via Roxygen2 or comments) to ensure it stays updated as the code changes.

### Specific Practices & Tools

### Roxygen2 for Function Documentation
For all custom functions in R, endeavor to use the `roxygen2` framework. This is the gold standard used by the tidyverse team. It allows us to generate formal documentation while keeping the descriptions immediately above the function definition.

**Key tags to use:**
* `@title`: A brief, one-line summary.
* `@description`: A more detailed explanation of what the function does.
* `@param`: Define each input argument (type, default value, and purpose).
* `@return`: Describe the output object.
* `@example`: Provide a minimal working example of the function in use.

```r
#' Calculate Shannon Diversity Index
#'
#' @description This function computes the Shannon-Wiener diversity index 
#' for a given vector of species counts.
#'
#' @param counts A numeric vector of species abundances.
#' @return A single numeric value representing the diversity index.
#' @export
#'
#' @examples
#' calc_shannon(c(10, 20, 5, 2))
calc_shannon <- function(counts) {
  p <- counts / sum(counts)
  return(-sum(p * log(p), na.rm = TRUE))
}
```

### In-line Comments
In-line comments should be used strategically to explain the logic that isn't immediately obvious from the code itself. We follow the [Tidyverse Style Guide](https://style.tidyverse.org/documentation.html) to keep our scripts clean and professional.

* **Focus on "Why," not "What":** Avoid redundant comments that simply restate the code (e.g., avoid `# filter for year 2020` followed by `filter(year == 2020)`). Instead, explain the reasoning: `# Excluding 2020 due to incomplete survey coverage in the Indo-Pacific`.
* **Formatting:** Always start a comment with `#` followed by a single space.
* **Section Breaks:** For long scripts, use four dashes after a comment to create foldable sections in RStudio and Positron to improve navigability: `# Data Cleaning ----`.
* **Pipe Documentation:** When using long `magrittr` or base R pipes (`%>%` or `|>`), place comments above the specific line of the pipe that performs a complex or non-intuitive transformation.
* **Todo Tags:** Use `# TODO:` to flag unfinished tasks or areas that require further optimization. This makes it easy for collaborators to search the codebase for pending items.

## AI-Assisted Documentation
We encourage the use of AI tools to accelerate the documentation process and improve the clarity of our codebase. However, AI is a co-pilot, not an autopilot; the researcher is ultimately responsible for the accuracy and technical integrity of all documentation.

### GitHub Copilot
GitHub Copilot is integrated into our IDEs (RStudio, Positron, and VS Code) to provide real-time suggestions.

* **Auto-completion:** As you begin writing a Roxygen2 block (`#' `), Copilot will often suggest `@param` descriptions and `@return` types based on the logic within the function body.
* **Inline Chat:** Use the shortcut `Cmd + I` (macOS) or `Ctrl + I` (Windows) to open the Copilot chat directly above a code block. You can use the `/doc` command to have Copilot generate a comprehensive documentation header for the highlighted code.
* **Unit Tests:** Use Copilot to generate the `@examples` section of your Roxygen2 headers, ensuring that the function's usage is well-documented and tested.

### Claude Code
Claude Code is our preferred tool for high-level project documentation, architectural overviews, and refactoring legacy scripts.

* **Project Summarization:** Claude is exceptionally skilled at reading entire directories. Use it to generate a comprehensive `README.md` for a new repository or to summarize the workflow of a complex analysis for a Quarto report.
* **Refactoring for Readability:** You can prompt Claude to audit your scripts: *"Refactor this R script to improve readability and add Tidyverse-style comments to any non-obvious data transformations."*
* **Style Conversion:** Use Claude to modernize documentation, such as converting unstructured Stata comments or old R scripts into standardized Roxygen2 format.

### Best Practices for AI Documentation
* **Verify Logic:** AI can "hallucinate" the intent of a function or describe a parameter incorrectly. Always read through AI-generated text to ensure it

## Documenting your analysis using Quarto notebooks

While code comments explain the "how," Quarto documents (`.qmd`) are used to explain the "so what." At emLab, Quarto is our primary tool for integrating code with scientific narrative, ensuring that our logic and results are inseparable.

* **Analytical Notebooks:** Use Quarto to interleave code chunks with visualizations and scientific interpretation. This format is required for internal peer reviews and sharing preliminary results with the lab.
* **Version Control:** Always commit both the `.qmd` source and the resulting `.md` or `.html` output (where appropriate) to GitHub to ensure collaborators can view rendered results without needing to re-run heavy computations.
* **Reproducible Reports:** Utilize Quarto’s ability to handle LaTeX for equations and BibTeX for citations to maintain academic rigor within our technical documentation.


## Package and Language Version Management

Reproducible research requires not just the same code, but the same computational environment. To ensure that our analyses can be replicated years later or shared across different machines today, emLab adheres to a tiered strategy for managing software versions. This "future-proofing" approach prevents the common frustration where code breaks because a package was updated or a system dependency changed.

### R

For R projects, we manage three distinct layers of the environment: the R version itself, the project-specific library of R packages, and the source of the packages. By managing each of these layers, we ensure that an update to your global R installation doesn't inadvertently break a legacy project.

#### R Version Management with `rig`

Before managing packages, we must manage R itself. Different versions of R can introduce breaking changes in the underlying C++ code or require different package binaries. Using the "latest" version of R is often fine for new work, but reproducing an analysis from three years ago often requires the exact R version used at that time.

We use [**rig** (The R Installation Manager)](https://github.com/r-lib/rig) to handle multiple R versions on a single machine seamlessly.

* **Why use rig?** It allows you to quickly switch between R versions without manual uninstalls. Crucially, it installs R in a way that doesn't require `sudo` privileges for library paths, preventing permissions issues that often plague multi-user servers. This ensures that `renv` is always drawing from the correct, isolated R binary.
* **Workflow:**
  * **Installation:** Install the version specified in the project documentation (e.g., `rig add 4.3.2`).
  * **Context Switching:** Use `rig default 4.3.2` to set the system-wide default, or use `rig run <version>` to start a specific session.
  * **Documentation:** Always document the R version used in the project's `README.md` and ensure it matches the `R` field in the `renv.lock` file.

#### Package Version Management with `renv`

Every emLab project should be "hermetic"—meaning its packages are isolated from the rest of your system. We use [**renv**](https://rstudio.github.io/renv/articles/renv.html) to create a private library of packages for each project.

* **Initialization:** Run `renv::init()` at the start of a project. This creates a local `.Rprofile` that tells R to use a project-specific library instead of your global one.
* **Daily Workflow:**
  * `renv::snapshot()`: Save the state of your library to the `renv.lock` file. This JSON file contains the exact version and source (CRAN, GitHub, Bioconductor) of every package. **Commit this file to GitHub.**
  * `renv::status()`: Frequently check if your local library is in sync with your lockfile.
  * `renv::restore()`: When pulling changes from GitHub, run this to sync your local library with the project's lockfile. This is the "magic button" that rebuilds the environment on a colleague's computer.
* **Efficiency:** `renv` uses a global cache. This means that if ten different projects use `ggplot2` version 3.4.0, it is only stored on your hard drive once, but linked into each project's library.
* **Collaboration:** Never include the `renv/library` folder in your git commits (it should be in `.gitignore`). Only commit the `renv.lock`, `.Rprofile`, and `renv/activate.R` files.

#### Posit Package Manager (PPM)

To ensure that `renv::restore()` always finds the exact same package versions and to speed up installation via pre-compiled binaries, we use [Posit Package Manager](https://packagemanager.posit.co/).

* **Fixed Snapshots:** Instead of pointing to the "latest" CRAN, which changes daily, we point to a specific "frozen" date. This eliminates the risk of "dependency hell" where Package A updates and is no longer compatible with Package B.
* **Binary Advantages:** PPM provides binaries for specific Linux distributions (like Ubuntu Jammy). This means that instead of waiting 20 minutes to compile a complex spatial package like `sf` or `terra` from source, you can download a pre-built version in seconds.

* **Configuration:** You can configure your IDE (RStudio or Positron) to use PPM by default for all new work so you don't have to manually call the options every time.

The Universal R Profile (Recommended)

In your R console, run the following to edit your `.Rprofile`: 

```r 
usethis::edit_r_profile()
```

Paste the following line and save:

```r
options(repos = c(PPM = "[https://packagemanager.posit.co/cran/__linux__/jammy/latest](https://packagemanager.posit.co/cran/__linux__/jammy/latest)"))
```

IDE-specific options

- *RStudio*: Go to Tools > Global Options... > Packages and paste the PPM URL into the "Primary CRAN repository" field.

- *Positron*: Open Settings (Cmd + ,), search for r.defaultRepositories, and add the CRAN/PPM URL to your `settings.json.`

**IMPORTANT**
Global vs. Project Settings: Global defaults are for your convenience. However, you must still include the specific repository URL in each project's `.Rprofile` to ensure reproducibility for your collaborators.

#### Common Pitfalls & Troubleshooting (R)

Using `rig` and `renv` together is robust, but keep an eye out for these common issues that can stall a project:

* **The R Version Mismatch:** You might open a project and find `renv` complaining that it was initialized with R 4.2.0 while you are currently running 4.4.1. 
    * **The Consequence:** R may try to install packages meant for a newer version into an older environment, causing "binary not found" errors.
    * **The Fix:** Use `rig run 4.2.0` (or set it as default) before opening the project. Always check your current R version with `version` or `R.version.string` if things feel "off."
* **Stale Lockfiles:** If you install a new package via `install.packages()` but forget to run `renv::snapshot()`, your `renv.lock` file won't reflect the change. 
    * **The Consequence:** When a collaborator tries to `renv::restore()`, they won't get the new package, and their code will break with "package not found."
    * **The Fix:** Make `renv::status()` a habit before you commit and push to GitHub. It will tell you if your lockfile and library are out of sync.
* **Missing System Dependencies:** `renv` manages R packages, but it does **not** manage system-level software like GDAL, GEOS, or PROJ (essential for emLab spatial work).
    * **The Consequence:** Package installation fails during the "compilation" phase with cryptic errors about missing headers or `.so` files.
    * **The Fix:** You must install the system library on your OS (e.g., via `brew install gdal` on Mac or `sudo apt install libgdal-dev` on Linux) before `renv` can successfully build the R package.
* **Committing the Library:** Occasionally, users accidentally add the `renv/library` folder to Git. 
    * **The Consequence:** This bloats the repository size by hundreds of megabytes and causes errors for others because those binaries are specific to your operating system and R version.
    * **The Fix:** Ensure your `.gitignore` includes `renv/library/`. If you've already committed it, you'll need to use `git rm -r --cached renv/library` to remove it from tracking.

### Python

For Python, we follow a similar isolation philosophy as we do with R to avoid "dependency drift."

#### Virtual Environments (`venv`)

Avoid installing packages to your "Global" Python, as this can break system-level tools. Always create a virtual environment within your project directory to keep dependencies contained.

* **Convention:** We use the name `.venv` for our environment folders.
* **Creation:** `python -m venv .venv`
* **Activation:**
  * Windows: `.venv\Scripts\activate`
  * Mac/Linux: `source .venv/bin/activate`
  * Once activated, your terminal prompt will usually change to show `(.venv)`, indicating that any `pip install` commands will only affect this project.
* **Git:** Ensure `.venv/` is added to your `.gitignore` to prevent committing thousands of small library files.

#### Dependency Tracking

To replicate a Python environment, we use `requirements.txt` for standard pip-based projects or `environment.yml` for more complex stacks.

* **The Importance of Pinning:** Simply listing `pandas` in a file is not enough; we should list `pandas==2.1.0`. This prevents "silent failures" where code runs but produces different numerical results due to underlying algorithm changes in newer package versions.
* **Exporting:** Use `pip freeze > requirements.txt` to capture every sub-dependency and its exact version.
* **Installing:** When joining a project, run `pip install -r requirements.txt` to recreate the environment instantly.

#### 3. Conda and Mamba

For projects with complex non-Python dependencies—specifically spatial libraries like GDAL, GEOS, or PROJ—standard pip often fails. In these cases, we recommend [**Mamba**](https://mamba.readthedocs.io/en/latest/).

* **Why Mamba?** Standard Conda can take hours to "solve" a complex environment (finding a set of versions that all work together). Mamba is a C++ implementation that does this in seconds.
* **Environment Files:** Use an `environment.yml` file to define both the Python version and the required packages from the `conda-forge` channel.
* **Documentation:** Document the creation command clearly:
  `mamba env create -f environment.yml`

#### Common Pitfalls & Troubleshooting (Python)

Python environment management is notoriously "leaky." Even with virtual environments, it is easy to accidentally run code in the wrong context. Watch out for these common emLab hurdles:

* **The "Shadow" Global Install:** You run `pip install` without realizing your virtual environment isn't active.
    * **The Consequence:** The package is installed to your system's global Python. Your code runs fine on your machine, but when you share the `requirements.txt` or `environment.yml` file, the package is missing, and your collaborator's code fails.
    * **The Fix:** Always check your terminal prompt for the `(.venv)` prefix. When in doubt, run `which python` (Mac/Linux) or `where python` (Windows) to ensure it points to your project folder, not `/usr/bin/python`.
* **Mixing Pip and Conda:** You use `pip install` inside a Mamba/Conda environment for a package that has complex C-dependencies.
    * **The Consequence:** This is the leading cause of "Environment Inconsistency" errors. Pip and Conda do not communicate well; Pip might overwrite a library that Conda relies on, leading to a broken environment that won't update.
    * **The Fix:** If you are using Mamba, **always** try `mamba install package_name` first. Only use `pip install` if the package is unavailable on `conda-forge`.
* **The GDAL/Spatial Nightmare:** Trying to install spatial libraries like `geopandas`, `fiona`, or `rasterio` via `pip`.
    * **The Consequence:** These packages require specific versions of system libraries (GDAL, PROJ, GEOS). Pip tries to compile these from source, which almost always fails on standard laptops without a massive headache.
    * **The Fix:** Use **Mamba** and an `environment.yml` file for any project involving spatial data. Mamba handles the system-level binaries so you don't have to.
* **Stale Requirements Files:** You've been working for weeks, installing new packages, but haven't updated your tracking file.
    * **The Consequence:** The `requirements.txt` file in your GitHub repo is months out of date. New lab members spend hours trying to debug "ModuleNotFoundError."
    * **The Fix:** Regularly run `pip freeze > requirements.txt` (for venv) or `mamba env export > environment.yml` (for Mamba) before every major push to GitHub.
* **Python Version Drift:** Your code uses f-strings or new syntax from Python 3.12, but your collaborator is on 3.9.
    * **The Consequence:** Syntax errors that look like code bugs but are actually version issues.
    * **The Fix:** Specifically define the Python version at the top of your `environment.yml` or in your `README.md`.

---

### Summary Table

| **Feature** | **R Strategy** | **Python Strategy** | **Primary Goal** |
| :--- | :--- | :--- | :--- |
| **Language Version** | `rig` | `pyenv` or `conda` | Isolate the interpreter from OS updates. |
| **Project Isolation** | `renv` | `venv` or `conda/mamba` | Prevent projects from conflicting with each other. |
| **Lockfile** | `renv.lock` | `requirements.txt` or `yml` | Record exact versions for 100% reproducibility. |
| **Binary Source** | Posit Package Manager | PyPI or Conda-Forge | Ensure fast, reliable installation of libraries. |

### External Resources & Further Reading

For a deeper dive into the tools and philosophies discussed in this SOP, we recommend the following resources:

#### R & Package Management
* [**rig: The R Installation Manager**](https://github.com/r-lib/rig): The official repository for `rig`. Includes detailed installation guides for macOS, Windows, and Linux, and advanced commands for managing Rtools.
* [**renv: Introduction to Project Environments**](https://rstudio.github.io/renv/articles/renv.html): The official "Get Started" guide for `renv`. It provides a clear overview of the package's philosophy and a comprehensive list of commands.
* [**Using Posit Package Manager (PPM)**](https://p3m.dev/): The public instance of PPM. Use the "Setup" button here to generate the exact repository URL for a specific date or operating system.
* [**CRAN Task View: Reproducible Research**](https://cran.r-project.org/view=ReproducibleResearch): A curated list of R packages and tools dedicated to making research more reproducible, from literate programming (Quarto/Knitr) to specialized environment managers.

#### Python & Environment Management
* [**Real Python: Virtual Environments Guide**](https://realpython.com/python-virtual-environments-a-primer/): An excellent, beginner-friendly primer on why virtual environments are necessary and how to use the built-in `venv` module.
* [**Mamba Documentation**](https://mamba.readthedocs.io/en/latest/): The user guide for Mamba. Essential reading if you are transitioning from standard Conda and want to understand why Mamba's "solver" is so much faster.
* [**Conda-Forge: The Community Repository**](https://conda-forge.org/docs/user/introduction.html): Since emLab relies on `conda-forge` for spatial libraries, this guide explains how the community maintains these packages and how to troubleshoot version conflicts.

## Pipeline Reproducibility

Modern research workflows often involve complex, interconnected scripts where the output of one file is the input for the next. Traditional "numbered script" approaches (e.g., `01\_clean.R`, `02\_model.R`) are prone to error: if you change a parameter in `01\_clean.R`, you must remember to manually re-run everything downstream.  
To ensure end-to-end reproducibility and efficiency, **emLab projects should use automated pipeline tools.** These tools track dependencies, skip tasks that are already up-to-date, and provide a clear map of the research workflow.

### The Philosophy of Pipeline Management

Adopting a pipeline tool is not just about learning a new package; it is a shift in research philosophy. We use these tools for several critical reasons:

* **Resilience to Change:** In research, data is rarely static. If a raw dataset is updated or a cleaning parameter is tweaked, manual workflows require the researcher to remember every downstream script affected. Automated pipelines handle this bookkeeping for you, ensuring the "state" of your results always matches the "state" of your code.  
* **Efficiency and "Lazy" Computing:** Many emLab analyses involve long-running operations (e.g., processing GFW data or training complex models). Pipeline tools are "lazy"—they only execute a task if its inputs or the code itself has changed. This saves hours of unnecessary computation and allows for rapid iteration on specific parts of the analysis.  
* **Trust and Transparency:** A pipeline tool serves as a "living" README. Instead of trusting a narrative description of how the analysis was run, a collaborator can look at a dependency graph and see exactly how every figure and table was derived.  
* **Collaborative Scalability:** In a team setting, pipelines reduce the "cognitive load" for new members. A new researcher can clone a repository and run a single command to see the entire project in action, rather than guessing which scripts are current and which are experimental leftovers.

### R Pipelines: The targets Package

For projects primarily using R, we highly recommend the [targets](https://books.ropensci.org/targets/) package. targets maintains a persistent cache and only re-runs the specific parts of your analysis that have changed.

#### Core Structure

A targets project typically follows this directory structure:  

```
project/  
├── _targets.R          \# The "instruction manual" for the pipeline  
├── R/                  \# Folder containing modular functions  
│   ├── functions.R       
├── data/               \# Raw data  
├── _targets/           \# (Automatic) Local cache/metadata  
└── report.qmd          \# Final output (e.g., paper figures and statistics)
```

#### Functionalize Your Code

Instead of long scripts, write small, modular functions in the `R/` directory. Each function should do one thing (e.g., "clean the sea surface temperature data").

#### Define the Pipeline (`_targets.R`)

This file defines the relationship between data, code, and outputs.  
\# \_targets.R  
```{r}
#| eval: false
library(targets)
source("R/functions.R")
tar_option_set(packages = c("tidyverse", "sf"))

list(
  # Track a raw data file
  tar_target(raw_data_file, "data/raw_samples.csv", format = "file"),

  # Clean data using a custom function from R/functions.R
  tar_target(data_cleaned, clean_my_data(raw_data_file)),

  # Run a model
  tar_target(model_results, run_analysis(data_cleaned)),

  # Render a Quarto report
  tar_quarto(final_report, "report.qmd")
)
```

#### Sharing the Data Store across Users

By default, targets stores its data in a local `_targets/` folder. Where you store this folder depends on the size of your objects:

* **Small Objects (\<50MB per file):** You can keep the `_targets/` folder in your GitHub repository. However, you must ensure you aren't committing large binary files that will bloat the repo history.  
* **Large Research Data:** For most emLab projects, analysis objects are too large for GitHub. In these cases, you should move the data store to emLab's shared infrastructure (Nextcloud GRIT storage).

##### Automating Path Detection

To ensure your pipeline works seamlessly across different team members' machines and our GRIT servers (quebracho or sequoia), use the following boilerplate at the top of your `_targets.R` file:  
```{r}
#| eval: false
# Determine base directory based on system/nodename  
data_directory_base <- ifelse(  
  Sys.info()["nodename"] %in% c("quebracho", "sequoia"),  
  "/home/emlab",  
  ifelse(  
    Sys.info()["sysname"] == "Darwin",  
    "/Users/Shared/nextcloud/emLab",  
    ifelse(  
      Sys.info()["sysname"] == "Windows",  
      "G:/Shared drives/nextcloud/emLab",  
      "/home/your_username/Nextcloud" \# Fallback for Linux users  
    )  
  )  
)

# Define project-specific directory  
project_directory <- glue::glue("{data_directory_base}/projects/current-projects/your-project-name")

# Set targets store to appropriate GRIT/Nextcloud directory  
tar_config_set(  
  project = "your_project_pipeline",  
  script = "_targets.R",  
  store = glue::glue("{project_directory}/data/_targets")  
)
```

**Crucially:** When using a shared store, only one person should run `tar_make()` at a time to avoid write conflicts in the metadata. Targets will not allow simeltaneous runs.

#### Multiple Pipelines in One Repo

Complex emLab projects often require split pipelines. For example:

1. **Pipeline A (Data Ingestion):** Pulls heavy data from BigQuery/GFW. Requires specific credentials and often runs on the emLab server/GRIT.  
2. **Pipeline B (Analysis):** Uses the pulled data for modeling. Accessible to all team members for local iteration.

To manage this, create two separate target files (e.g., `_targets_data.R` and `_targets_analysis.R`) and a `_targets.yaml` file in the root directory to define them:  

```{r}
#| eval: false
# _targets.yaml
data\_pipeline:  
  script: \_targets\_data.R  
  store: \_targets\_data\_store  
analysis\_pipeline:  
  script: \_targets\_analysis.R  
  store: \_targets\_analysis\_store
```

You can then run them specifically from the console:  
`tar_make(script = "_targets_data.R", store = "_targets_data_store")`

#### Rendering Quarto Books

If your final output is a Quarto Book rather than a single document, use the `tar_quarto_book()` function from the tarchetypes package. This should be the final target in your list.  

```{r}
#| eval: false
# _targets.R
library(targets)
library(tarchetypes)

list(
  # ... previous targets ...
  tar_quarto_book(website, path = ".")
)
```

### Non-R Pipelines: GNU Make

For projects using Python, Stata, or a mix of languages, we recommend [GNU Make](https://www.gnu.org/software/make/). Make uses a Makefile to define "rules" for creating files.

#### The Makefile Syntax

A rule consists of a **target** (the file you want to create), **prerequisites** (files it depends on), and a **recipe** (the command to run).  
target: prerequisites  
	recipe

#### Example Workflow (Python/Stata)

```{bash}
#| eval: false
# Makefile

# Run everything
all: results/paper.pdf

# Clean raw data (Python)
data/processed.csv: scripts/clean_data.py data/raw.csv
	python scripts/clean_data.py

# Run analysis (Stata)
results/model_output.log: scripts/analysis.do data/processed.csv
	stata -batch do scripts/analysis.do

# Render report (Quarto)
results/paper.pdf: paper.qmd results/model_output.log
	quarto render paper.qmd --to pdf
```

#### How to use Make

* **Run the pipeline:** Simply type make in your terminal. Make will check the timestamps of all files and only run the steps where the prerequisites are newer than the target.  
* **Dry run:** Type make \-n to see what commands *would* be run without actually executing them.

### Minimum Requirement: README Documentation

If a project does not yet use an automated tool like targets or Make, it is **mandatory** to provide a detailed "How to Run" section in the repository's README.md. At a absolute minimum, this must include:

1. **Script Execution Order:** A numbered list of scripts in the exact order they must be run.  
2. **Dependencies:** Which data files are required for each script and which outputs they produce.  
3. **Environment Setup:** Instructions on how to set up the necessary libraries or software versions (e.g., using renv or conda).

### Best Practices for All Pipelines

1. **Documentation is Mandatory:** No matter what pipeline or tool is used, the repository README.md must clearly document how to execute the analysis. A new collaborator should be able to clone the repo and know exactly which command to run (e.g., "Run tar\_make() in the R console" or "Type make in the terminal").  
2. **Atomic Steps:** Each step in the pipeline should be small enough to be easily understood but large enough to justify being its own "target."  
3. **File Tracking:** Always include the script itself as a prerequisite/dependency. If you change the code, the output should be considered "outdated."  
4. **No Manual Interventions:** A pipeline is only reproducible if it can be run from start to finish on a new machine without a human clicking "save," "export," or manually moving files in the middle.

## Internal code review

Peer-review of each other's code is a great method for mutual learning through exposure to different coding styles and packages, ensuring reproducibility, catching any bugs, and uncovering opportunities for doing things better. We recognize that this takes time, but we believe that the time commitment is worth it for us to all become better coders and write better code. As a best practice, we recommend that for any project that has at least two researchers, there should be a systematic review of any major coding elements by the researcher who did not initially write the code. By following the best practices outlined in the `Coding Pipeline` section of this SOP, the original coder can ensure that the reviewing coder knows exactly which pieces of code are critical for their review.

Code reviews should be respectful and collegial in manner. The goal is to make all of our work better and make all us better coders, not to criticize or single anyone out.

The original coder and reviewing coder should coordinate on the best method for feedback, whether that is through direct pull requests to the code, an email outlining suggestions, or a Google doc outlining suggestions.

For further guidance on effective code reviewing techniques and guiding principles, we recommend using the [Tidyteam code review principles](https://tidyverse.github.io/code-review/). These were developed by the `tidyteam` for maintaining packages such as those found in the `tidyverse` and `tidymodels`. These guidelines are based on Google's [Code Review Developer Guide](https://google.github.io/eng-practices/review/), another helpful resource.

## Exploratory Data Analysis

Much of the work emLab does is Exploratory Data Analysis (EDA), especially during the beginning stages of a project when we are familiarizing ourselves we new datasets and rapidly prototyping data wrangling and modeling approaches. The following quote from [Hadley Wickham's R for Data Science](https://r4ds.had.co.nz/exploratory-data-analysis.html) describes EDA as:

> EDA is an iterative cycle. You: 1) Generate questions about your data; 2) Search for answers by visualising, transforming, and modelling your data; 3 Use what you learn to refine your questions and/or generate new questions.

EDA is a creative and exploratory process without defined rules. Hadley Wickham's chapter on [EDA](https://r4ds.had.co.nz/exploratory-data-analysis.html) is an excellent place to start for ideas, but we encourage creativity and exploration during this phase of a project.

However, we recommend as a best practice that all EDA should include an "Interpretations, questions, and new ideas" section. The researcher doing EDA is poised in an excellent position for providing initial interpretations of the data, raising outstanding questions about the data, and generating novel insights and potential research questions. This section should contain the following information:

* Interpretations about what you found. Does it match our intuition? Is there anything surprising?
* Questions about the data. Are there any questions about how to use or interpret the data? Are there any potential problems with the data or analysis?
* New insights and research questions. Have you uncovered any new insights about the data or research? Have you discovered any new research questions that might be worth pursuing?

By including a section in each EDA analysis that discusses these aspects, we can build insight generation into our workflows. This information should be in an easy to find location on the Team Drive, such as in a `markdown-reports` directory that contains EDA PDFs generated by R Markdown. This way, other members of the project team can read it, digest it, and respond with further ideas.
