# Exploratory data analysis

Exploratory data analysis (EDA) is where you identify data quality issues and refine hypotheses. While EDA can be "messy," at emLab we treat it as a formal part of the research record. Using Quarto for EDA allows us to document not just what we found, but what we looked for and why certain paths were abandoned.

## The "notebook" philosophy

Use .qmd files as an interactive laboratory notebook. Unlike a final report, an EDA document should include:

* **Data Sanity Checks:** Summaries of missing values, range checks, and distribution plots.  
* **Dead Ends:** If an approach didn't work, keep the code but explain why in the text. This prevents future collaborators (or your future self) from repeating the same mistake.  
* **Plot annotations:** Every plot should be followed by a short sentence describing the takeaway (e.g., *"The outlier in 2014 appears to be a recording error in the raw sensor data"*).

## Leveraging code folding

EDA often requires long blocks of code to generate diagnostic plots. To keep the document readable while remaining transparent, use code-fold at the chunk level.  

```{r}
#| label: eda-missing-data  
#| code-fold: true  
#| code-summary: "Show code for missingness analysis"  
# Complex visualization code here...
```

## Interactive exploration tools

Static tables are often insufficient for exploring large datasets. We recommend using interactive widgets within your Quarto EDA documents (rendered to HTML) to allow for deeper inspection.

### Interactive tables with DT

Use the DT package to create searchable, sortable tables. This is invaluable for spotting specific outliers or checking metadata.  
```{r}
#| label: eda-inspect-data  
library(DT)
mtcars |>
  datatable(options = list(pageLength = 10, autoWidth = TRUE))
```

### Quick summaries with `skimr`

Instead of standard `summary()`, use `skimr::skim()` to get a high-level overview of distributions and missingness directly in your Quarto output.  

```{r}
#| label: eda-skim  
mtcars |>
  skimr::skim()
```

## Organizing EDA documents

For large projects, do not crowd a single file with all exploratory work. Instead:

1. **Create a dedicated eda/ folder** in your repository.  
2. **Use a naming convention** like eda-01-initial-cleaning.qmd, eda-02-spatial-joins.qmd, etc.  
3. **If using a Quarto Book**, include these under a "Research Notebook" or "Appendices" section to keep the main narrative focused.

## Best practices for EDA

* **eval: true / echo: true**: Ensure code is visible and executed so others can reproduce your exploratory steps.  
* **Avoid View()**: Never leave View(df) in a code chunk, as it will break the rendering process. Use head() or DT::datatable() instead.  
* **Self-Correction**: If you find an error in the data during EDA, document the fix in the text and move the correction logic to your data cleaning script/pipeline node.
