# Functional programming {#sec-functional-programming}

Functional programming is a style of writing code that helps make emLab models and software easy to
**understand,
combine,
improve,
and debug.**

Functional programming is not a separate language or package — it is a way of structuring code, based on **applying functions to data**. A rudimentary example of a function in the programming sense is^[This section uses R code in [Quarto](../04-quarto/index.qmd) chunks, hence the `[1]` prefix.]

```{r}
#| label: sum-fun
#| message: false
sum(1,1) # <1>
```

A functional programming approach focuses on building *functions* involving several or many lines of code, which may then be used in various contexts throughout a given project. It can be thought of as describing a **method**. It contrasts with *imperative* programming, which involves *scripts*, where each line gives a computer a **command**. For comparison:

::: {.panel-tabset}
## Functional

```{r}
say_its <- function(sun) {
  if (sun == "up") return("Good morning")
  if (sun == "down") return("Good night")
}

say_its("up")
say_its("down")
```

## Imperative

```{r}
sun <- "up"
if (sun == "up") print("Good morning")
if (sun == "down") print("Good night")

sun <- "down"
if (sun == "up") print("Good morning")
if (sun == "down") print("Good night")
```
:::

## Don't Repeat Yourself

The guiding instinct of functional programming is the **DRY principle** — *Don't Repeat Yourself*. When the same logic is copied across several scripts, every fix or improvement has to be made in every copy, and any copy that gets missed becomes a silent bug. Writing that logic once, as a function, leaves a single place to change it.

A useful rule of thumb comes from Hadley Wickham in [*R for Data Science*](https://r4ds.hadley.nz/functions):

> You should consider writing a function whenever you've copied and pasted a block of code more than twice (i.e. you now have three copies of the same code).

DRY is also why functional programming is a prerequisite for the pipeline tools described in [Pipeline reproducibility](pipeline-reproducibility.qmd): both `targets` and GNU Make chain together small, well-named functions or scripts as the nodes of the dependency graph. The cleaner your functions, the cleaner the pipeline.

## Combining functions

R is a multi-paradigm language, meaning it can work with either style of programming. However, emLab data scientists rely on R's many thousands of *packages*. When you add an R package to your library, you can then build your code using functions written by others.

```{r}
#| label: tibble
library(tibble)

tibble(
  small = letters[1:3], 
  big = LETTERS[1:3]
)
```

The functional programming approach would be to build their functions into your own, likely incorporating other functions^[Base (a.k.a. vanilla) R includes lots of functions geared towards math and statistics even before you add anything to its library, such as `sqrt()` and `matrix()`] too:

```{r}
#| label: combine
#| message: false
#| warning: false
library(stringr)

sentence_square <- function(sentence) {
  words = str_split_1(sentence, pattern = " ") # from `stringr`
  dimension = ceiling(sqrt(length(words)))
  square = matrix(words, nrow = dimension)
  return(as_tibble(square))
}

sentence_square("Wherefore art thou Romeo")
```

This is an example of nesting, where calling `sentence_square()` indirectly calls the other functions used in its recipe. Part of what makes functional programming such a useful approach is that the arguments^[comma-separated items inside the parentheses] of a function are accessible to functions nested within it: 

```{r}
#| label: fp-nesting-2
pluralize <- function(word) {
  paste0(word, "s")
}

pluralize("apple")

group <- function(item, count) {
  paste(count, pluralize(item))
}

group("orange", 5)
```

Here, `pluralize()` takes its `word` argument from the `item` argument of `group()`. A variable defined within a function is said to be **local**, and a function's arguments are among its local variables. Variables that are local to a function are not accessible in a broader scope:

```{r}
#| label: error
#| error: true
kitchen <- function(light = "on") {
  stove = "gas"  
  return(mget(ls())) # describe local variables
}

light
stove
kitchen()
```

## Improving functions

Combining functions lets you build new behavior; **wrapping** lets you adjust behavior that already exists. A wrapper is a function whose main job is to call one external function while supplying your preferred defaults or an extra step. The original package is left untouched — you have simply given your project its own, improved version.

A frequent emLab use is enforcing a consistent look for figures. [`ggplot2`](https://ggplot2.tidyverse.org/) draws plots by adding layers together, but its defaults are generic. Wrapping the styling layers once means every figure in a project matches, without copying the same code into every script:

```{r}
#| label: fp-wrap-ggplot
#| message: false
#| fig-height: 3
library(ggplot2)

emlab_style <- function(plot) {
  plot +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

mpg_plot <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "#2aa198") +
  labs(
    title = "Heavier cars travel fewer miles per gallon",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  )

emlab_style(mpg_plot)
```

The real payoff is maintainability: if the lab later decides bold titles were a mistake, that change happens in `emlab_style()` alone, rather than in every script that draws a figure.^[As custom functions accumulate, consider bundling them into an R package of their own — emLab encourages sharing project tooling this way, especially alongside a publication.]

## Debugging functions

Given input it cannot handle, a function can do one of two things: fail loudly, or carry on and return something wrong. emLab's [guiding principle of accuracy](index.qmd) warns that the most costly errors are the *silent* ones — code that runs cleanly and produces plausible output while doing the wrong thing. Functions that check their own inputs turn those silent errors into loud, immediate ones.

R provides three signals for communicating from inside a function, in increasing severity:

* `message()` — an informational note, such as a progress update. Execution continues.
* `warning()` — something looks wrong but is recoverable. Execution continues, with a flag.
* `stop()` — a fatal problem. Execution halts before bad output can spread downstream.

```{r}
#| label: fp-debug-stop
#| error: true
greet <- function(sun) {
  if (!sun %in% c("up", "down")) {
    stop('`sun` must be "up" or "down", not "', sun, '".')
  }
  if (sun == "up") return("Good morning")
  if (sun == "down") return("Good night")
}

greet("sideways")
```

Because `greet()` validates its input, the mistake surfaces *right here*, with a message naming the problem — rather than several functions deeper, disguised as a strange result. `message()` and `warning()` handle the less severe cases:

```{r}
#| label: fp-debug-warn
mean_positive <- function(x) {
  if (any(x < 0)) warning("Negative values dropped before averaging.")
  message("Averaging ", sum(x >= 0), " of ", length(x), " values.")
  mean(x[x >= 0])
}

mean_positive(c(4, 8, -2, 10))
```

These signals help whoever *runs* a function. Readability helps whoever *reads* it — at emLab, usually a collaborator or your future self. The logic inside a function deserves the same care as any script: descriptive names, consistent style, and comments that explain *why*. See [Code styling](code-styling.qmd) and [Code documentation](code-documentation.qmd) for the details.

Additionally, if you are actively trying to debug a function, adding a call for `browser()` inside the function can be an incredibly powerful temporary tool for understanding what's going on inside your function. By inserting `browser()` into a function, whenever you run the function it will pause exactly where the call is located. This will allow you to see exactly what is going on in your session, including the values stored in objects in the function's environment. In the case below, we insert a `browser()` call after calculting the mean value. When you run this function, it will drop you right into the function at that exact location. You can then inspect `mean_value` to ensure it matches your expectation.

```{r}
#| label: fp-debug-browser
mean_positive <- function(x) {
  if (any(x < 0)) warning("Negative values dropped before averaging.")
  message("Averaging ", sum(x >= 0), " of ", length(x), " values.")
  mean_value <- mean(x[x >= 0])

  browser()
  mean_value
}

mean_positive(c(4, 8, -2, 10))
```

## Repetitive functions

### Prefer `apply`/`map` over `for` loops when iterating over a collection

When the repetition you need is "do this same thing to every element of a vector or list," reach for the **apply/map** family of functions before a `for` loop. In base R that means `lapply()`, `sapply()`, or `vapply()`; in the tidyverse, `purrr::map()` and its typed variants (`map_dbl()`, `map_chr()`, `map_lgl()`, ...).

```{r}
#| label: apply-vs-for
words <- c("apple", "banana", "cherry")

# Imperative: a for loop that mutates an output vector
lengths_loop <- integer(length(words))
for (i in seq_along(words)) {
  lengths_loop[i] <- nchar(words[i])
}
lengths_loop

# Functional: applying nchar() across the vector using purrr
# purrr::map() returns a list, which can sometimes be helpful
purrr::map(words, nchar)

# Other purrr::map_*() functions return different types of objects if you want a different output.
# In this case, we might want an integer vector, so we use can map_int().
purrr::map_int(words, nchar)

# Functional: applying nchar() across the vector using sapply()
sapply(words, nchar)
```

The functional form is a single expression that takes a collection in and gives a collection out, without making any other changes to the environment in which it runs.

## Exceptions

emLab strives for a functional programming mindset, but here are examples of cases where scripting is sometimes used.

### Recursive functions

R supports recursive functions, which call themselves:

```{r}
#| label: fp-recursion
factorial <- function(n) {
  if (n <= 1) return(1)    # base case: stop here
  n * factorial(n - 1)   # recursive case: shrink toward the base case
}

factorial(4)
```

However, emLab code prioritizes readability, and recursion is a case where many find the imperative equivalent more intuitive:

```{r}
#| label: while-loop
n = 4
factorial = 1

while (n > 1) {
  factorial = factorial * n
  n = n - 1
}

factorial
```

Notably, one strong argument for the functional programming paradigm is that someone else probably already made a function that does what you want to do. If you have a good sense of the tooling available and its applicability to your use case, you may sometimes have to write hardly any code at all. 

Using the product `prod()` function from base R:

```{r}
n = 4 
factorial = prod(1:n) # R parses `a:b` as the integers between a and b
factorial
```

### Pipes

When composing functions, e.g., `f(g(x))`, there are sometimes cases with many intermediate states. The two ways to address this while sticking strictly to a functional programming paradigm are nesting functions:

```{r}
chef_it_up <- function(eggs) {
  omelette <- plate(fold(fry(crack(eggs))))
  return(omelette)
}
```

or using intermediate **local** variables^[which aren't retrievable outside of the function]:

```{r}
chef_it_up <- function(eggs) {
  goop <- crack(eggs)
  circle <- fry(goop)
  burrito <- fold(circle)
  omelette <- plate(burrito)
  return(omelette)
}
```

Using pipes `|>` to link up functions changes the state of an invisible variable multiple times, meaning it draws from the imperative programming paradigm. However, it makes code more readable than either purely functional approach, while also using computer memory efficiently by overwriting each step's input with its output.

```{r}
chef_it_up <- function(eggs) {
  omelette <- eggs |>
    crack() |>
    fry() |>
    fold() |>
    plate()
  return(omelette)
}
```

It is worth mentioning that pipes pass each step's results as the first argument of the next function. This tends to work out alright when functions are designed according to the best practice of putting the main variable first. It is also possible to use a placeholder `_` so a pipe goes to a different argument:

```{r}
hit <- function(pitcher = "Who", batter = "What") {
  if (pitcher == "Babe Ruth") return("Strikeout!")
  if (batter == "Babe Ruth") return("Home run!")
}

first_up = "Babe Ruth"

first_up |>
  hit() # Babe Ruth as pitcher

first_up |>
  hit(batter = _) # Babe Ruth as batter

```

---

# Recommended readings

* Wickham, Hadley. 2019. *Advanced R*, 2nd ed. Chapman & Hall/CRC. See the chapters on "Functions" and "Conditions," and the "Functional programming" section. <https://adv-r.hadley.nz/>.

* Wickham, Hadley, Mine Çetinkaya-Rundel, and Garrett Grolemund. 2023. *R for Data Science*, 2nd ed. O'Reilly. See the "Functions" chapter. <https://r4ds.hadley.nz/functions>.

* Bryan, Jenny. 2020. *Object of type 'closure' is not subsettable: Debugging in R*. This is an excellent talk, slides, and GitHub repo going over best practices for debugging in R. <https://github.com/jennybc/debugging>.

