# Functional programming {#sec-functional-programming}

Functional programming (FP) is a style of writing code that emphasizes
**applying functions to data** rather than mutating variables step by step. It
is not a separate language or package — it is a way of structuring R code that
makes analyses easier to read, test, and reproduce.

This chapter is a hands-on tutorial. Every R code block below is executed when
the SOP website is rendered in our GitHub Actions pipeline, so the output you
see is generated live from the code shown.

## Why functional programming?

Most emLab analyses involve doing the *same thing* to *many things*: cleaning
each of a dozen data files, fitting a model for each region, or summarizing
every species in a dataset. Functional programming gives us a vocabulary for
these patterns that is:

* **Readable** — `map(files, read_csv)` states intent more clearly than a
  `for` loop that grows a list.
* **Reliable** — small, pure functions can be tested in isolation, which
  supports our [guiding principle of accuracy](index.qmd).
* **Reproducible** — code without hidden state behaves the same on every
  machine and every run.

We rely on base R's functional tools and on
[**purrr**](https://purrr.tidyverse.org/), the tidyverse iteration package.

```{r}
#| label: fp-setup
#| message: false
library(purrr)
library(dplyr)
```

## Functions are values

The foundational idea of FP is that **functions are ordinary values**. You can
store a function in a variable, put it in a list, or pass it as an argument to
another function — exactly as you would a number or a data frame.

```{r}
#| label: fp-functions-as-values
# Assign a function to a variable
celsius_to_f <- function(temp_c) {
  temp_c * 9 / 5 + 32
}

# Store several functions in a list
converters <- list(
  to_fahrenheit = celsius_to_f,
  to_kelvin = function(temp_c) temp_c + 273.15
)

converters$to_kelvin(20)
```

A function that takes another function as an argument (or returns one) is
called a **higher-order function**. `map()`, which we meet below, is the
higher-order function you will use most.

## Write small, pure functions

A **pure function** depends only on its inputs and produces only a return
value — it reads no global variables and causes no side effects (no writing
files, no modifying objects elsewhere, no printing). Pure functions are the
building blocks of reliable analysis code because their behavior is fully
determined by their arguments.

```{r}
#| label: fp-pure-function
# Pure: output depends only on the arguments
standardize <- function(x) {
  (x - mean(x)) / sd(x)
}

round(standardize(c(10, 12, 14, 16, 18)), 2)
```

Compare that to an **impure** version that reaches outside itself for data:

```{r}
#| label: fp-impure-function
#| eval: false
# Impure: silently depends on a global `reference_mean`.
# Avoid this — the result changes if `reference_mean` changes elsewhere.
center_bad <- function(x) {
  x - reference_mean
}
```

::: {.callout-tip}
## Guideline
If a function needs a value, pass it in as an argument. Functions that pull
from the global environment are the source of bugs that are invisible until
the script is run in a different order or on someone else's machine.
:::

## Iterate with `map()` instead of loops

The `map()` family replaces most `for` loops. `map(.x, .f)` applies the
function `.f` to every element of `.x` and **always returns a list** of the
same length.

```{r}
#| label: fp-map-basic
numbers <- list(1, 2, 3, 4)

map(numbers, \(x) x^2)
```

Note the `\(x) x^2` syntax: this is R's **anonymous function** shorthand
(R 4.1+), equivalent to `function(x) x^2`. purrr also accepts a formula
shorthand, `~ .x^2`, where `.x` stands for the current element.

### Type-stable variants

Returning a list is safe but often you want a plain vector. The typed
variants — `map_dbl()`, `map_int()`, `map_chr()`, `map_lgl()` — return an
atomic vector and **error if the function returns the wrong type**. That
strictness catches mistakes early.

```{r}
#| label: fp-map-typed
# Average mpg by number of cylinders, returned as a named numeric vector
cyl_groups <- split(mtcars, mtcars$cyl)

map_dbl(cyl_groups, \(df) mean(df$mpg))
```

```{r}
#| label: fp-map-chr
# Build a character summary for each group
map_chr(cyl_groups, \(df) paste0(nrow(df), " cars"))
```

### Iterating over two or more inputs

`map2()` walks two vectors in parallel; `pmap()` handles any number of inputs
supplied as a list (or the columns of a data frame).

```{r}
#| label: fp-map2
weights <- c(0.2, 0.3, 0.5)
values <- c(100, 200, 300)

map2_dbl(weights, values, \(w, v) w * v)
```

```{r}
#| label: fp-pmap
params <- tibble::tibble(
  base = c(2, 3, 4),
  exponent = c(3, 2, 1)
)

pmap_dbl(params, \(base, exponent) base^exponent)
```

### When you only want side effects

If you are iterating purely for a side effect — printing, saving plots,
writing files — use `walk()`. It runs the function for each element and
returns the input invisibly, signalling "this is for the side effect, not the
value."

```{r}
#| label: fp-walk
walk(c("Cleaning step 1", "Cleaning step 2"), \(msg) cat(msg, "\n"))
```

## Combining results with `reduce()`

`reduce()` collapses a vector or list into a single value by applying a
two-argument function cumulatively. It is the functional way to express
"combine all of these."

```{r}
#| label: fp-reduce
# Sum a list of numbers
reduce(list(1, 2, 3, 4, 5), `+`)
```

A common emLab use is joining several data frames into one:

```{r}
#| label: fp-reduce-join
band_tables <- list(
  dplyr::band_members,
  dplyr::band_instruments
)

reduce(band_tables, dplyr::full_join, by = "name")
```

`accumulate()` is the same idea but keeps every intermediate result, which is
useful for inspecting a running total or a cumulative process.

```{r}
#| label: fp-accumulate
accumulate(1:5, `+`)
```

## Functional programming inside the tidyverse

These tools shine in routine data work. A frequent pattern is reading and
row-binding many files with `map()` — here simulated with in-memory data so the
chunk runs without external files:

```{r}
#| label: fp-map-dfr
# Pretend each element is a file path; the function returns one data frame
fake_files <- c("site_a", "site_b", "site_c")

read_one <- function(site) {
  tibble::tibble(site = site, reading = nchar(site) * 10)
}

map(fake_files, read_one) |>
  list_rbind()
```

`map()` also pairs naturally with `dplyr` for column-wise operations. Inside
`across()`, you supply a function just as you would to `map()`:

```{r}
#| label: fp-across
mtcars |>
  group_by(cyl) |>
  summarise(across(c(mpg, hp, wt), \(x) round(mean(x), 1)))
```

## Predicate functions: `keep()`, `discard()`, `detect()`

A **predicate** is a function that returns a single `TRUE` or `FALSE`. purrr
provides higher-order functions that take a predicate to filter or search.

```{r}
#| label: fp-predicates
nums <- list(4, 7, 10, 3, 12, 1)

keep(nums, \(x) x > 5)      # elements passing the predicate
discard(nums, \(x) x > 5)   # elements failing the predicate
detect(nums, \(x) x > 5)    # first element passing the predicate
```

## Best practices

* **Prefer `map()` over `for` loops** when building a result. Reach for a loop
  only when each iteration genuinely depends on the previous one and
  `reduce()`/`accumulate()` do not fit.
* **Use the typed `map_*()` variants** (`map_dbl()`, `map_chr()`, …) whenever
  you expect a vector — the type check is free error-catching.
* **Keep functions small and pure.** One function, one job, no hidden inputs.
* **Name your functions for what they return**, in `snake_case`, following our
  [code styling guide](code-styling.qmd).
* **Don't over-abstract.** If a piece of logic appears once, a plain
  expression is clearer than a function. Refactor to a function on the second
  or third repetition.
* **Test pure functions in isolation.** Because their output depends only on
  their inputs, a handful of `stopifnot()` checks or testthat tests fully
  pin down their behavior.

## Recommended readings

* Wickham, Hadley. 2019. *Advanced R*, 2nd ed. Chapman & Hall/CRC. Chapters on
  "Functionals" and "Function factories." <https://adv-r.hadley.nz/>

* Wickham, Hadley, Mine Çetinkaya-Rundel, and Garrett Grolemund. 2023.
  *R for Data Science*, 2nd ed. O'Reilly. Chapter 26, "Iteration."
  <https://r4ds.hadley.nz/iteration>

* purrr package documentation and cheatsheet.
  <https://purrr.tidyverse.org/>
