# Pipeline reproducibility {#sec-pipeline-reproducibility}

Modern research workflows often involve complex, interconnected scripts where the output of one file is the input for the next. Traditional sequential approaches (e.g., first running `01_clean.R`, then running `02_model.R`) are prone to error: if you change a parameter in `01_clean.R`, you must remember to manually re-run everything downstream.  

To ensure end-to-end reproducibility and efficiency, **emLab projects should use automated pipeline tools.** These tools track dependencies, skip tasks that are already up-to-date, and provide a clear map of the research workflow.

## The philosophy of pipeline management

These tools are useful for several practical reasons:

* **Automatic dependency tracking:** In research, data is rarely static. If a raw dataset is updated or a cleaning parameter is tweaked, you would otherwise need to remember every downstream script affected. Automated pipelines track these dependencies automatically.
* **Skipping up-to-date steps:** Many emLab analyses involve long-running operations (e.g., processing GFW data or training complex models). Pipeline tools are "lazy"—they only execute a task if its inputs or the code itself has changed. This saves significant computation time during iteration.
* **A clear map of the analysis:** A pipeline's dependency graph shows exactly how every figure and table was derived, which is more reliable than a written description of the run order.
* **Easier onboarding:** In a team setting, new team members can run a single command to execute the full pipeline, rather than guessing which scripts are current and which are experimental leftovers.

## R pipelines: the targets package

For projects primarily using R, we highly recommend the [targets](https://books.ropensci.org/targets/) package. targets maintains a persistent cache and only re-runs the specific parts of your analysis that have changed.

### Targets repo structure

A targets project typically follows this directory structure:  

```
project/  
├── _targets.R          # The "instruction manual" for the pipeline  
├── R/                  # Folder containing modular functions  
│   ├── functions.R       
├── data/               # Raw data  
├── _targets/           # (Automatic) Local cache/metadata  
└── report.qmd          # Final output (e.g., paper figures and statistics)
```

### Functionalize your code

Instead of long scripts, write small, modular functions in the `R/` directory. Each function should do one thing (e.g., "clean the sea surface temperature data").

The `_targets.R` file defines the relationship between data, code, and outputs.  

```{r}
#| eval: false
library(targets)
source("R/functions.R")
tar_option_set(packages = c("tidyverse", "sf"))

list(
  # Track a raw data file
  tar_target(raw_data_file, "data/raw_samples.csv", format = "file"),

  # Clean data using a custom function from R/functions.R
  tar_target(data_cleaned, clean_my_data(raw_data_file)),

  # Run a model
  tar_target(model_results, run_analysis(data_cleaned)),

  # Render a Quarto report
  tarchetypes::tar_quarto(final_report, "report.qmd")
)
```

### Sharing the data store across users

By default, targets stores its data in a local `_targets/` folder. Where you store this folder depends on the size of your objects:

* **Small objects (<50MB per file):** You can keep the `_targets/` folder in your GitHub repository. However, you must ensure you aren't committing large binary files that will bloat the repo history.  
* **Large data:** For most emLab projects, analysis objects are too large for GitHub. In these cases, you should move the data store to emLab's shared infrastructure (Nextcloud GRIT storage).

#### Automating path detection

To ensure your pipeline works seamlessly across different team members' machines and our GRIT servers (quebracho or sequoia), use the following boilerplate at the top of your `_targets.R` file:  
```{r}
#| eval: false
# Determine base directory based on system/nodename
data_directory_base <- ifelse(
  Sys.info()["nodename"] %in% c("quebracho", "sequoia"),
  "/home/emlab",
  ifelse(
    Sys.info()["sysname"] == "Darwin",
    "/Users/Shared/nextcloud/emLab",
    ifelse(
      Sys.info()["sysname"] == "Windows",
      "G:/Shared drives/nextcloud/emLab",
      "/home/your_username/Nextcloud" # Fallback for Linux users
    )
  )
)

# Define project-specific directory
project_directory <- glue::glue(
  "{data_directory_base}/projects/current-projects/your-project-name"
)

# Set targets store to appropriate GRIT/Nextcloud directory
tar_config_set(
  project = "your_project_pipeline",
  script = "_targets.R",
  store = glue::glue("{project_directory}/data/_targets")
)
```


::: {.callout-important}

When using a shared store, only one person should run `tar_make()` at a time to avoid write conflicts in the metadata. Targets will not allow simultaneous runs.

:::


### Multiple pipelines in one repo

Complex emLab projects often require split pipelines. For example:

1. **Pipeline A (Data ingestion):** Pulls heavy data from BigQuery/GFW. Requires specific credentials and often runs on the emLab server/GRIT.  
2. **Pipeline B (Analysis):** Uses the pulled data for modeling. Accessible to all team members for local iteration.

To manage this, create two separate target files (e.g., `_targets_data.R` and `_targets_analysis.R`) and a `_targets.yaml` file in the root directory to define them:  

```yaml
# _targets.yaml
data_pipeline:  
  script: _targets_data.R  
  store: _targets_data_store  
analysis_pipeline:  
  script: _targets_analysis.R  
  store: _targets_analysis_store
```

You can then run them specifically from the console:  
`tar_make(script = "_targets_data.R", store = "_targets_data_store")`

## Rendering Quarto documents in a `targets` pipeline

When your `targets` pipeline produces a Quarto document, the render step should be the last target in your pipeline — and how you define it matters. The right tool for this is `tarchetypes::tar_quarto()`.

### Why not just use `targets::tar_target()`?

It might seem natural to call the Quarto render function inside a regular target:

```{r}
#| eval: false
# Not recommended
targets::tar_target(report, quarto::quarto_render("analysis/final_report.qmd"))
```

The issue is that `targets` sees this as an opaque function call with no connection to the rest of your pipeline - it does not know which targets are refereced inside the quarto doc. It has no way of knowing which upstream targets the document reads, so it cannot tell when the document is out-of-date. In practice, this means you either re-render unnecessarily on every `tar_make()` run, or — worse — end up with a stale report that `targets` believes is current.

### Use `tarchetypes::tar_quarto()` instead

`tarchetypes::tar_quarto()` fixes this by scanning your `.qmd` file for `targets::tar_read()` and `targets::tar_load()`, so that it can recognize any target objects that are called as explicit upstream dependencies. The document is then re-rendered only when one of those dependencies has actually changed.

```{r}
#| eval: false
# Recommended
tarchetypes::tar_quarto(
  name = quarto_notebook,
  path = "qmd/quarto_notebook.qmd",
  quiet = FALSE
)
```

Inside your `.qmd` file, use `targets::tar_read(target_name)` or `targets::tar_load(target_name)` to access pipeline results. These are the calls that `tarchetypes::tar_quarto()` scans to build the dependency graph, so make sure all pipeline inputs are accessed this way rather than loaded through other means.

Using `targets::tar_target()` directly is technically possible, but you would need to manually declare all dependencies yourself to get correct invalidation behavior. We therefore strongly recommend `tarchetypes::tar_quarto()` for most use cases.

## Non-R pipelines: GNU Make

For projects using Python, Stata, or a mix of languages, we recommend [GNU Make](https://www.gnu.org/software/make/). Make is language-agnostic, comes pre-installed on macOS and most Linux distributions (and is easy to install on Windows via WSL), and uses a single plain-text `Makefile` to describe how each output in the pipeline is built.

Unlike `targets`, which tracks dependencies by hashing R objects, Make reasons in terms of **files and timestamps**: a target is considered out-of-date whenever any of its prerequisites has a newer modification time. This is a coarser signal than `targets` provides — Make cannot tell whether a file's *contents* actually changed, only that it was touched — but it is enough for most multi-language pipelines and is easy to reason about.

### The Makefile syntax

A `Makefile` is a list of **rules**. Each rule has three parts:

```{.makefile}
target: prerequisites
<TAB>recipe
```

* **target** — the file the rule produces (e.g. `data/processed.csv`).
* **prerequisites** — files the target depends on. If any prerequisite is newer than the target, the recipe runs.
* **recipe** — the shell command(s) Make runs to build the target. The indentation must be a literal **tab character**, not spaces. Silent space-for-tab substitution by editors is the single most common Makefile bug; configure your editor to preserve tabs in your `Makefile`.

### Example workflow (Python/Stata)

```{.makefile filename="Makefile"}
.PHONY: all clean

# Default target: build everything
all: results/paper.pdf

# Clean raw data (Python)
data/processed.csv: scripts/clean_data.py data/raw.csv
	python scripts/clean_data.py

# Run analysis (Stata)
results/model_output.log: scripts/analysis.do data/processed.csv
	stata -batch do scripts/analysis.do

# Render report (Quarto)
results/paper.pdf: paper.qmd results/model_output.log
	quarto render paper.qmd --to pdf

# Wipe generated outputs so the pipeline can be re-run from scratch
clean:
	rm -f data/processed.csv results/model_output.log results/paper.pdf
```

A few things worth pointing out in this example:

* `.PHONY: all clean` tells Make that `all` and `clean` are command names, not files. Without it, a stray file named `all` in the project root would make Make think the pipeline is already built and refuse to do anything.
* `all` is listed first, so a bare `make` (no arguments) builds it. By convention, `all` depends on the final outputs of the pipeline.
* Each script appears as a prerequisite *and* in its own recipe. This way, edits to `scripts/clean_data.py` invalidate `data/processed.csv` and trigger a rerun on the next `make`.

### How to use Make

| Command | Effect |
|---|---|
| `make` | Build the default target (the first non-phony target in the file, by convention `all`). |
| `make <target>` | Build a specific target, e.g. `make results/paper.pdf`. |
| `make -n` | Dry run — print the commands Make *would* execute, without running them. Useful for sanity-checking a pipeline change. |
| `make -B` | Force-rebuild every target, ignoring timestamps. |
| `make -j N` | Run up to `N` independent steps in parallel. |
| `make clean` | Run the `clean` rule above (or whichever cleanup rule the project defines). |

### Pitfalls and tips

* **Always declare phony targets.** Any rule that does not actually produce a file with the target's name (`all`, `clean`, `lint`, `test`, ...) belongs in `.PHONY`.
* **Include the script itself as a prerequisite.** Otherwise edits to the analysis code won't trigger a rebuild and downstream outputs will go stale silently.
* **For scripts that don't produce a tracked file** (e.g. a Stata script that only prints to console), redirect the output to a sentinel file (`results/model_output.log` in the example above) so Make has something to time-stamp.
* **Document the entry point in the README.** A first-time collaborator should be able to clone the repo, type `make`, and have the whole pipeline build itself.

## Minimum requirement: README documentation

If a project does not yet use an automated tool like targets or Make, it is **mandatory** to provide a detailed "How to Run" section in the repository's README.md. At a absolute minimum, this must include:

1. **Script execution order:** A numbered list of scripts in the exact order they must be run.  
2. **Dependencies:** Which data files are required for each script and which outputs they produce.  
3. **Environment setup:** Instructions on how to set up the necessary libraries or software versions (e.g., using renv or conda).

## Best practices for all pipelines

1. **Documentation is mandatory:** No matter what pipeline or tool is used, the repository README.md must clearly document how to execute the analysis. A new collaborator should be able to clone the repo and know exactly which command to run (e.g., "Run tar_make() in the R console" or "Type make in the terminal").  
2. **Atomic steps:** Each step in the pipeline should be small enough to be easily understood but large enough to justify being its own "target."  
3. **File tracking:** Always include the script itself as a prerequisite/dependency. If you change the code, the output should be considered "outdated."  
4. **No manual interventions:** A pipeline is only reproducible if it can be run from start to finish on a new machine without a human clicking "save," "export," or manually moving files in the middle.
